---
title: "Calculating $p$-values in R: A Quick and Dirty Guide"
author: "Jeffrey M. Pisklak"
date: "`r Sys.Date()`"
output: html_document
---

<style>
details > summary {
  padding: 4px;
  cursor: pointer;
  font-size: 1.5em;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(RColorBrewer)
# palette.colors(palette = "Okabe-Ito")
crit_col <- brewer.pal(8, 'Dark2')[4]
prob_col <- brewer.pal(8, 'Dark2')[2]
alpha <- 0.05
df <- 4
```

- Critical Region ($\alpha = .05$) is shown in light red. 
- Solid vertical black line is the test-statistic.
- The probability of a test-statistic (i.e. $p$-value) is shown in brown.
- Degrees of freedom are set to 4 in all examples.
- Additional notes pertaining to one- and two-sided confidence interval calculations are also provided.

<br>

# **Two-Tailed Tests**

<h5>
- For two-tailed tests, what is matters is whether the test-statistic lands in the left or right tail; i.e., whether it is negative or positive.
- Note that the probability (i.e., $p$-value's area under the curve) is mirrored in the case of two-tailed tests.
</h5>

### 1. If your **test-statistic is negative** (e.g., $T = -1.5$), then you are going to use . . .

```{r echo = TRUE}
pt(-1.5, df = 4) * 2
```

```{r echo = FALSE, fig.width = 5, fig.height = 4, message = FALSE}
t_stat <- -1.5
p <- pt(t_stat, df = df) * 2
x <- seq(from = -6, to = 6, by = .001)
dens <- dt(x, df = df)
dist_df <- data.frame(x, dens)

sub_title <- paste("E.g., T = ", t_stat, ", df = ", df, sep = "")
plt_anno <- paste("T = ", t_stat, "\np = ", round(p, 3),sep = "")

ggplot(dist_df, aes(x = x, y = dens)) +

  # Title
  labs(
    title = "Two-Tailed",
    subtitle = sub_title
  ) +

  # Left tail probability
  geom_ribbon(
    data = subset(dist_df, x < -1.5),
    aes(ymax = dens),
    ymin = 0,
    fill = prob_col,
    colour = NA,
    alpha = 0.5
  ) +

  # Right tail probability
  geom_ribbon(
    data = subset(dist_df, x >= 1.5),
    aes(ymax = dens),
    ymin = 0,
    fill = prob_col,
    colour = NA,
    alpha = 0.5
  ) +

  # Test-stat vertical line
  geom_vline(
    xintercept = t_stat,
    colour = "black",
    linewidth = 1
  ) +
  annotate("text",
    x = -3,
    y = .25,
    label = plt_anno
  ) +

  # Axis Scale
  scale_x_continuous(
    breaks = seq(-4, 4, 1)
  ) +
  
  coord_cartesian(xlim = c(-4, 4)) +

  # Left tail 95% threshold
  geom_ribbon(
    data = subset(
      dist_df,
      x < qt(alpha / 2, df = df)
    ),
    aes(ymax = dens),
    ymin = 0,
    fill = crit_col,
    colour = NA,
    alpha = 0.5
  ) +

  # Right tail 95% threshold
  geom_ribbon(
    data = subset(
      dist_df,
      x > qt(1 - alpha / 2, df = df)
    ),
    aes(ymax = dens),
    ymin = 0,
    fill = crit_col,
    colour = NA,
    alpha = 0.5
  ) +
  geom_line(linewidth = 1) +
  xlab("t-score") +
  ylab("Density") +
  theme(
    axis.text.x = element_text(colour = "black", size = 12),
    axis.title = element_text(colour = "black", size = 16),
    axis.ticks.y = element_blank(),
    axis.text.y = element_blank()
  )
```

<br>

<details>
  <summary>
  Confidence Interval Information
  </summary>

<br>

$\begin{aligned}
\text{Confidence Interval} &= \bar{x} \pm \text{(critical $t$-score)} \cdot \text{(standard error)} \\
&= [\text{ lower bound}, \text{ upper bound }]
\end{aligned}$

```{r}
# Critical t-calculation
alpha <- 0.05
abs(qt(alpha / 2, df = 4))
```

- The absolute value (`abs()`) is used because the confidence interval equation already accounts for the direction of each boundary through addition (+) and subtraction (-). This eliminates the need to compute separate critical $t$-scores for each side.

</details>

<br>

<br>

### 2. If your **test-statistic is positive** (e.g., $T = 1.5$), then you are going to use . . .

```{r echo = TRUE}
pt(1.5, df = 4, lower.tail = FALSE) * 2
```


```{r echo = FALSE, fig.width = 5, fig.height = 4, message = FALSE}
t_stat <- 1.5
p <- pt(1.5, df = df, lower.tail = FALSE) * 2
x <- seq(from = -6, to = 6, by = .001)
dens <- dt(x, df = df)
dist_df <- data.frame(x, dens)

sub_title <- paste("E.g., T = ", t_stat, ", df = ", df, sep = "")
plt_anno <- paste("T = ", t_stat, "\np = ", round(p, 3),sep = "")

ggplot(dist_df, aes(x = x, y = dens)) +

  # Title
  labs(
    title = "Two-Tailed",
    subtitle = sub_title
  ) +

  # Left tail probability
  geom_ribbon(
    data = subset(dist_df, x < -1.5),
    aes(ymax = dens),
    ymin = 0,
    fill = prob_col,
    colour = NA,
    alpha = .5
  ) +

  # Right tail probability
  geom_ribbon(
    data = subset(dist_df, x > 1.5),
    aes(ymax = dens),
    ymin = 0,
    fill = prob_col,
    colour = NA,
    alpha = .5
  ) +
  
  # Test-stat vertical line
  geom_vline(
    xintercept = 1.5,
    colour = "black",
    linewidth = 1
  ) +
  annotate("text",
    x = 3,
    y = .25,
    label = plt_anno
  ) +

  # Axis Scale
  scale_x_continuous(
    breaks = seq(-4, 4, 1)
  ) +
  
  coord_cartesian(xlim = c(-4, 4)) +

  # Left tail 95% threshold
  geom_ribbon(
    data = subset(
      dist_df,
      x < qt(alpha / 2, df = df)
    ),
    aes(ymax = dens),
    ymin = 0,
    fill = crit_col,
    colour = NA,
    alpha = 0.5
  ) +

  # Right tail 95% threshold
  geom_ribbon(
    data = subset(
      dist_df,
      x > qt(1 - alpha / 2, df = df)
    ),
    aes(ymax = dens),
    ymin = 0,
    fill = crit_col,
    colour = NA,
    alpha = 0.5
  ) +
  geom_line(linewidth = 1) +
  xlab("t-score") +
  ylab("Density") +
  theme(
    axis.text.x = element_text(colour = "black", size = 12),
    axis.title = element_text(colour = "black", size = 16),
    axis.ticks.y = element_blank(),
    axis.text.y = element_blank()
  )
```

<br>

<details>
  <summary>
  Confidence Interval Information
  </summary>

<br>

$\begin{aligned}
\text{Confidence Interval} &= \bar{x} \pm \text{(critical $t$-score)} \cdot \text{(standard error)} \\
&= [\text{ lower bound}, \text{ upper bound }]
\end{aligned}$

```{r}
# Critical t-calculation
alpha <- 0.05
abs(qt(alpha / 2, df = 4))
```

- The absolute value (`abs()`) is used because the confidence interval equation already accounts for the direction of each boundary through addition (+) and subtraction (-). This eliminates the need to compute separate critical $t$-scores for each side.

</details>

<br>

<br>

# **One-tailed tests**

<h5>
- For one-tailed tests, what is matters is where the critical region (red) is located (left or right tail). This is determined by the null hypothesis.
- Whether the test-statistic is positive or negative has no bearing on how the $p$-value is calculated.
</h5>

### 3. If your **critical region is in the left tail** (e.g., $H_0 : \mu \geq 1.5$) then you are going to use . . .

```{r echo = TRUE}
pt(1.5, df = 4)
```

```{r echo = FALSE, fig.width = 5, fig.height = 4, message = FALSE}
t_stat <- 1.5
p <- pt(1.5, df = df)
x <- seq(from = -6, to = 6, by = .001)
dens <- dt(x, df = df)
dist_df <- data.frame(x, dens)

sub_title <- paste("E.g., T = ", t_stat, ", df = ", df, sep = "")
plt_anno <- paste("T = ", t_stat, "\np = ", round(p, 3),sep = "")

ggplot(dist_df, aes(x = x, y = dens)) +

  # Title
  labs(
    title = "One-Tailed",
    subtitle = sub_title
  ) +

  # Left tail probability
  geom_ribbon(
    data = subset(dist_df, x < 1.5),
    aes(ymax = dens),
    ymin = 0,
    fill = prob_col,
    colour = NA,
    alpha = .5
  ) +
  
  # Test-stat vertical line
  geom_vline(
    xintercept = 1.5,
    colour = "black",
    linewidth = 1
  ) +
  annotate("text",
    x = 3,
    y = .25,
    label = plt_anno
  ) +

  # Axis Scale
  scale_x_continuous(
    breaks = seq(-4, 4, 1)
  ) +
  
  coord_cartesian(xlim = c(-4, 4)) +

  # Left tail 95% threshold
  geom_ribbon(
    data = subset(
      dist_df,
      x < qt(alpha, df = df)
    ),
    aes(ymax = dens),
    ymin = 0,
    fill = crit_col,
    colour = NA,
    alpha = 0.5
  ) +
  geom_line(linewidth = 1) +
  xlab("t-score") +
  ylab("Density") +
  theme(
    axis.text.x = element_text(colour = "black", size = 12),
    axis.title = element_text(colour = "black", size = 16),
    axis.ticks.y = element_blank(),
    axis.text.y = element_blank()
  )
```

<br>

<details>
  <summary>
  Confidence Interval Information
  </summary>

<br>

$\begin{aligned}
\text{Confidence Interval} &= \bar{x} + \text{(critical $t$-score)} \cdot \text{(standard error)} \\
&= (-\infty, \text{ upper bound }]
\end{aligned}$

```{r}
# Critical t-calculation
alpha <- 0.05
abs(qt(alpha, df = 4))
```

- The confidence interval is expressed in terms of the alternative hypothesis $H_1: \mu < 1.5$.
- For one-tailed tests, the critical $t$-score is calculated using the *entire* significance level $\alpha$.
- The absolute value (`abs()`) is used because the direction of the boundary is already specified by the addition (+) in the confidence interval equation.

</details>

<br>

<br>

### 4. If your **critical region is in the right tail** (e.g., $H_0 : \mu \leq 1.5$) then you are going to use . . .

```{r echo = TRUE}
pt(1.5, df = 4, lower.tail = FALSE)
```

```{r echo = FALSE, fig.width = 5, fig.height = 4, message = FALSE}
t_stat <- 1.5
p <- pt(t_stat, df = df, lower.tail = FALSE)
x <- seq(from = -6, to = 6, by = .001)
dens <- dt(x, df = df)
dist_df <- data.frame(x, dens)

sub_title <- paste("E.g., T = ", t_stat, ", df = ", df, sep = "")
plt_anno <- paste("T = ", t_stat, "\np = ", round(p, 3),sep = "")

ggplot(dist_df, aes(x = x, y = dens)) +

  # Title
  labs(
    title = "One-Tailed",
    subtitle = sub_title
  ) +

  # Right tail probability
  geom_ribbon(
    data = subset(dist_df, x > 1.5),
    aes(ymax = dens),
    ymin = 0,
    fill = prob_col,
    colour = NA,
    alpha = .5
  ) +
  
  # Test-stat vertical line
  geom_vline(
    xintercept = 1.5,
    colour = "black",
    linewidth = 1
  ) +
  annotate("text",
    x = 3,
    y = .25,
    label = plt_anno
  ) +

  # Axis Scale
  scale_x_continuous(
    breaks = seq(-4, 4, 1)
  ) +
  
  coord_cartesian(xlim = c(-4, 4)) +

  # Right tail 95% threshold
  geom_ribbon(
    data = subset(
      dist_df,
      x > qt(1 - alpha, df = df)
    ),
    aes(ymax = dens),
    ymin = 0,
    fill = crit_col,
    colour = NA,
    alpha = 0.5
  ) +
  geom_line(linewidth = 1) +
  xlab("t-score") +
  ylab("Density") +
  theme(
    axis.text.x = element_text(colour = "black", size = 12),
    axis.title = element_text(colour = "black", size = 16),
    axis.ticks.y = element_blank(),
    axis.text.y = element_blank()
  )
```

<br>

<br>

<details>
  <summary>
  Confidence Interval Information
  </summary>

<br>

$\begin{aligned}
\text{Confidence Interval} &= \bar{x} - \text{(critical $t$-score)} \cdot \text{(standard error)} \\
&= [\text{ lower bound}, \infty)
\end{aligned}$

```{r}
# Critical t-calculation
alpha <- 0.05
abs(qt(alpha, df = 4))
```

- The confidence interval is expressed in terms of the alternative hypothesis $H_1: \mu > 1.5$.
- For one-tailed tests, the critical $t$-score is calculated using the *entire* significance level $\alpha$.
- The absolute value (`abs()`) is used because the direction of the boundary is already specified by the subtraction (-) in the confidence interval equation.

</details>

<br>

<br>